Comparing probabilistic accounts of probability judgments

### 2021-09-07 paper-rmd/paper-papaja.Rmd
added figs

---
title             : "Untitled probability judgment paper draft"
shorttitle        : "DRAFT"

@@

abstract: |
  Bayesian theories of cognitive science are centered around the idea that much of cognition is probabilistic. However, people’s explicit probability judgments often incoherently violate the laws of probability. Two recent proposals, the “Probability Theory plus Noise” [@costello.watts2014] and “Bayesian Sampler” [@zhu.etal2020] theories of probability judgments, both seek to account for these biases while maintaining that mental credences are fundamentally probabilistic. The two theories diverge significantly in their treatments of conditional probabilities, but comparing these theories poses challenges in accounting for model complexity. Here, I cast these theories into a Bayesian data analysis framework that supports principled model comparison through the $LOO_{ic}$ information criteria and compared the fits of both models on data collected by Zhu and colleagues (2020). My findings support the Bayesian Sampler’s account of conditional probability judgments, and I find the data are best explained by a “hybrid” model that can be seen as a modification of both theories.

@@

Bayesian theories and approaches offer a unified formal framework for cognitive science [@tenenbaum.etal2011] with remarkable successes across domains, including perception [e.g. @kersten.etal2004], memory [@anderson1991], language [@xu.tenenbaum2007], and reasoning [@lu.etal2012]. At the heart of the Bayesian project is the idea that cognition is fundamentally probabilistic: that people reason according to subjective degrees of belief which follow the laws of probability and, in particular, that they are revised in light of evidence according to Bayes' Rule. It is somewhat embarrassing then, that one area where these theories seem to fall down is in describing human "beliefs" of the simple and everyday sort, such as beliefs like "it will rain tomorrow", “vaccines are safe,” or “this politician is trustworthy”.

@@

Unfortunately, people’s explicit probability judgments routinely violate the axioms of probability theory. For example, human probability judgments often exhibit the “conjunction fallacy”: people will often judge the conjunction of two events (e.g. “Tom Brady likes football and miniature horses”) as being more probable than one of the events in isolation (e.g. “Tom Brady likes miniature horses”), a plain and flagrant violation of probability theory [@tversky.kahneman1983]. Other demonstrations of the incoherence of probability judgments include disjunction fallacies, subadditivity or “unpacking” effects [@tversky.koehler1994], and a variety of other effects illustrating the incoherence of human probability judgments [for an accessible review, see @kahneman2013]. Altogether these findings have led many researchers to abandon the notion that degrees of belief are represented as probabilities.

Recently however, two groups of researchers have proposed theories of human probability judgments that account for biases and apparent incoherence in these judgments while maintaining that mental credences are fundamentally probabilistic [@costello.watts2014; @zhu.etal2020]. Both of these theories build on the increasingly popular notion that a variety of human reasoning tasks are accomplished by drawing a limited number of samples from probabilistic mental models [see also @chater.etal2020; @dasgupta.etal2017].

@@

Costello & Watts [-@costello.watts2014; -@costello.watts2016; -@costello.watts2018] proposed a theory of probability judgment they call the “Probability Theory plus Noise” theory (PT+N). In the PT+N model, mental “samples” are drawn from a probabilistic mental model of events and are then “read” with noise, so that some positive examples will be read as negative and some negative read as positive. This results in probability judgments that reflect probabilistic credences perturbed by noise. Under the simplest form of the PT+N model, the expected value of probability judgments is:

@@

where *d* represents the probability with which samples will be misread, or the amount of noise in the judgment process (where by assumption a maximum of 50% of samples can be misread, so that d is a number in the range [0, .50]). The PT+N theory provides a unified accounts for a wide variety of biases in probability judgment that were previously attributed to different types of heuristics [@costello.watts2014; @costello.watts2016; @costello.watts2017; @costello.watts2018].

Meanwhile, Zhu, Sanborn, & Chater [-@zhu.etal2020] have proposed a Bayesian model of probability judgment they call the “Bayesian Sampler.” Under this model, probability judgment is itself seen as a process of Bayesian inference. To judge the probability of an event, a limited number of samples are again drawn from a mental model of the event. Then, those “observed” samples are integrated with a prior over probabilities to produce a probability judgment. The effect of this prior is to regularize the judgments, pushing them away from extreme values. They argue this regularization of judgments is adaptive, in that it pushes judgments and decisions away from extremes that might, e.g. encourage taking large risks based on a small number of mental samples. However, this hedging comes at the cost of systematic incoherence and biases. Like the PT+N model, their proposed model accounts for a wide array of biases in probability judgments, including the novel biases identified by Costello and Watts [@costello.watts2014; @costello.watts2016].

Under the simplest version of the Bayesian Sampler model, the expected value of a probability judgment is given by the following formula:

@@

where *N* represents the number of mental samples drawn. The mentally-observed samples are integrated with a Bayesian prior over probabilities, represented with a symmetric beta distribution parameterized by $\beta$. Zhu and colleagues [-@zhu.etal2020] show that the *N* and $\beta$ parameters of their model can be related to the *d* parameter of the PT+N model via the following bridging formula:

@@

@@

By explaining the incoherence of human probability judgments using coherent mental probabilities, both models have the potential to rescue the larger project of Bayesian cognitive science as applied to everyday beliefs [@chater.etal2020]. However, because they diverge in their treatment of conditional probability judgments, the two theories fit very differently into the larger Bayesian project. Bayesian cognitive theories are fundamentally theories of inductive reasoning: Bayes' rule describes how existing beliefs should be updated conditional on the observation of different kinds of evidence. So, the conditioning of beliefs is at the heart of these theories.

@@

$$ P_e(A|B) = \frac{(1-2d)^2P(A \land B) + d(1-2d)\big(P(A)+P(B)\big)+d^2}{(1-2d)P(B)+d} $$

@@

Zhu, Sanborn, and Chater (2020) compared their Bayesian Sampler model against Costello & Watts’ (2014; 2016; 2017; 2018) PT+N model as explanations for human probability judgments in two experiments. Unfortunately, their results were somewhat equivocal. They fit both “simple” and “complex” versions of each model and computed a Bayesian Information Criteria score (BIC) for all models. The complex versions of these models introduce additional parameters d’ and N’ that allow for different patterns of judgments for conjunctive and disjunctive judgments as compared with simple probability judgments. These additional parameters are needed in order to explain conjunction fallacies [@costello.watts2017; @zhu.etal2020] . Table 1 below presents the total BIC scores computed for each model as originally fit, using the authors’ original code and saved model outputs [@zhu.etal2020, Supplementary materials].

@@

In both experiments, the simple version of the Bayesian Sampler scores best, but the complex version of the PT+N model comes in second (lower BIC scores are better). It is not obvious what conclusions should be drawn from these results. Of course, the simple Bayesian Sampler model appears to win by the numbers. But conjunction fallacies are an extremely robust empirical finding [@mellers.etal2001; @sides.etal2002] and are clearly present in the data collected by Zhu and colleagues (2020). So we might rule out the simple variants of the models on the grounds that they will fail to capture important qualitative features of the data, which would instead favor the PT+N theory. For their part, Zhu and colleagues (2020) elect to split the baby by averaging the simple and complex model scores together. This approach somewhat favors the Bayesian Sampler theory overall, though they are appropriately cautious about drawing any strong conclusions.

@@

When comparing and selecting models based on their predictive accuracy, it is necessary to correct for the potential for these models to “overfit” the data. Typically this correction comes in the form of a “complexity penalty,” penalizing models in proportion to their flexibility for accommodating different patterns of data [@gelman.etal2014]. Zhu and colleagues (2020) warn that BIC, which penalizes models based only on the number of parameters in the model, cannot fully account for the differences in the models’ complexity.

There are at least three challenges to accounting for model complexity in the comparison of the PT+N and Bayesian Sampler models. First, the models differ not only in the number of parameters but in the domain of those parameters. Zhu and colleagues (2020) restrict the range of the $\beta$ parameter in their Bayesian Sampler model to fall within the interval [0,1]. This restriction constrains the Bayesian Sampler model’s prior distribution to the class of “ignorance priors”. Via the bridging conditions (eq X), they show that this restricts the “noise level” for the Bayesian Sampler,  represented in implied d under the PT+N model, to [0, $1/3$] whereas the PT+N model permits noise values between [0, 0.50].

Second, it is not immediately clear how the models’ structural differences impact their flexibility. To be sure, Zhu and colleagues’ (2020) bridging condition makes it clear that the PT+N model is more flexible when it comes to predicting unconditional probabilities. But, what impact does the PT+N model’s treatment of conditional probabilities have on model complexity? Is this component of the model a sort of Ptolemaic epicycle, adding complexity to the model that should be penalized? Or, does it constitute a commitment to novel predictions that thereby constrain its flexibility? Determining model complexity a priori is not always straightforward when the models being compared differ structurally.

Finally, one potential explanation for the relative weakness of the complex model versions is Zhu and colleagues’ (2020) use of “unpooled” models, where parameters are estimated independently for each individual participant. Adding a parameter for each participant may effectively over-penalize the complex variants relative to the simple variants, especially if there is actually limited heterogeneity across participants. In contrast, a comparison of fully pooled variants of the simple and complex models (where a single population parameter is used for all participants) would require adding only one extra parameter to the penalty term. Partial pooling is a solution that balances between these extreme approaches, allowing for an accounting of heterogeneity without over-penalizing in cases where heterogeneity is low.

@@

Here, I cast both the Bayesian Sampler and PT+N models into a Bayesian data analysis framework that permits a more decisive (or at least alternative) comparison. A Bayesian analysis allows issues of model complexity to be addressed through comparisons of model fit based on information criteria, such as $WAIC$ and $LOO_{ic}$. These indices approximate leave-one-out cross validation, thereby offering a measure of the ability of a model to capture new unseen data. Importantly, these indices also allow for complexity penalties that directly capture differences in the flexibility of the models, rather than being estimated based solely on the number of distinct parameters [@gelman.etal2014; @vehtari.etal2017]. In addition, this framework also supports straightforward implementation of hierarchical versions of these models with partial pooling. This allows for information about model parameters to be shared across participants, resulting in an overall reduction in model complexity and a more realistic test of the models.

@@

Zhu, Sanborn, & Chater (2020) conducted two experiments to compare the PT+N and Bayesian sampler theories. These experiments asked participants to judge the probability of different events in various combinations. Following prior work by Costello and Watts [e.g. -@costello.watts2016; -@costello.watts2018], both experiments focused on the everyday events of different kinds of weather.

Experiment 1 asked about the events [icy, frosty] and [normal, typical] (e.g. “what is the probability that the weather in London is normal and not typical?”).  The authors’ goal was to ask about highly correlated events, but the events used are perhaps nearly perfectly correlated. Because the terms used to describe these events are nearly synonymous, there is a concern about the interpretation of the statements evaluated in this experiment. This is especially clear, as the authors note, for disjunctive trials such as “normal or typical,” where “or typical” might not be read as a disjunction but rather an elaborative clause.  In light of these concerns, I excluded the disjunctive trials from Experiment 1 from my analyses.

Experiment 2 focused on more moderately correlated events, [cold, rainy] and [windy, cloudy], that do not admit these misinterpretations. In addition, a third experimental condition asking about [warm, snowy] was also included in the experiment, but was dropped from the analyses reported in the paper. Exploring the raw responses from this condition reveals a substantial fraction of “zero” and “one” responses for certain trials. This may reflect a different response process than was intended. For instance, some participants may have engaged in deductive reasoning to judge that it is not possible for the weather to at once be warm and snowy, and therefore responded with zero—failing to properly consider that it is possible (at least logically) for it to be warm and snowy at different times within the same day. Given these potentially aberrant responses, I followed Zhu and colleagues (2020) in ignoring data from this condition.

@@

The PT+N model defines probability judgments as

@@

In contrast, the Bayesian Sampler model defines probability judgments as

@@

Fixing d and d’ or N and N’ equal yields the “simple” variant of each of the models, which treat conjunctive and disjunctive probability judgments identically to simple probability judgments.

Notice that for each model the probability judgments depend on underlying subjective probabilities, derived from a mental sampling process. These subjective probabilities are unobserved, and must be estimated as a latent variable. Here, they are estimated as a four-dimensional dirichlet distribution for each subject, representing the probability of the elementary events ($A \land B$, $\neg A \land B$, $A \land \neg B$, $\neg A \land \neg B$).

Zhu, Sanborn & Chater (2020) implement completely unpooled models with separate d, d’, N, N’, and $\beta$ parameters for each participant. Translated into a Bayesian Framework, we can write the PT+N model as:

\begin{figure}
  \begin{minipage}{.5\textwidth}
    \flushright
    \input{ptn_platediagram}
  \end{minipage}%
  \begin{minipage}{.5\textwidth}
    \flushleft
    \small
    \begin{align*}
      y_{ij} &\sim beta\big(\mu_{ij} \kappa, (1-\mu_{ij}) \kappa \big) \\
      \mu_{ij} &= f_{PT+N}(\overrightarrow{p_i}, d_j, d_j)  \\
      d_j &= \frac{expit(\delta_j)}{2} \\
      d_j’ &= \frac{expit\big(\delta_j + \exp(\Delta\delta_j)\big)}{2} \\
      \overrightarrow{p_i} &\sim Dirichlet(\overrightarrow{1}) \\
      \delta_j &\sim Normal(0,1) \\
      \Delta\delta_j &\sim Normal(0,1) \\
      \kappa &\sim HalfCauchy(20)
    \end{align*}
  \end{minipage}
  \caption{Complex PT+N model diagram and formula specifications.}
\end{figure}

@@

Recall that Zhu and colleagues (2020) identified a bridging condition relating b and N in the Bayesian Sampler model to the d parameter of the PT+N model. To support direct comparisons of the models, I parameterize the Bayesian Sampler model according to the implied d and d’, rather than directly according to its $\beta$, N, and N’ parameters. I constrain d to [0, $1/3$] for the Bayesian Sampler model to reflect the assumption that $\beta \in [0, 1]$. This allows the same priors to be used for the corresponding Bayesian Sampler and PT+N models, simplifying their comparison.

The Bayesian Sampler model is therefore identical to the PT+N model save for the following changes to $\mu_{ijk}$, $d$, and $d'$, as shown below:

\begin{align*}
  \mu_{ij} &= f_{BS}(\overrightarrow{p_i}, d_j, d_j)  \\
  d_j &= \frac{expit(\delta_j)}{3} \\
  d_j’ &= \frac{expit\big(\delta_j + \exp(\Delta\delta_j)\big)}{3} \\
\end{align*}

@@

Both of these models can also be implemented as hierarchical models with partial pooling for the d and d’ parameters (implicitly, for N and N’ in the case of the Bayesian Sampler). This should prevent over-penalizing complex versions of the model, as well as potentially improving their overall predictive fit. The hierarchical implementation adds parameters for the population-level d and d’ as well as a parameter controlling the standard deviation of the distribution for the subject-level effects. For ease of interpretation, the centered parameterization is shown below, although the actual models used a non-centered parameterization to improve sampling efficiency [@papaspiliopoulos.etal2007].

\begin{figure}
  \begin{minipage}{.5\textwidth}
    \flushright
    \input{bs_platediagram_mlm}
  \end{minipage}%
  \begin{minipage}{.5\textwidth}
    \flushleft
    \small
    \input{bs_equation_mlm}
  \end{minipage}
  \caption{Hierarchical complex Bayesian Sampler model diagram and formula specifications.}
\end{figure}

Finally, I also explored fitting versions of the Bayesian Sampler model that allowed values of $\beta > 1$. Restricting $\beta$ to [0,1] restricts the prior distribution of the Bayesian sampler to the class of “ignorance priors” [@zhu.etal2020]. However, it is also possible that people bring informative priors to the probability judgment task. Indeed, Zhu and colleagues (2020) acknowledge there are situations where an informative prior may be warranted [see e.g., @fennell.baddeley2012]. If $\beta$ is unrestricted, allowed to fall in the domain [0, $\infty$] then the Bayesian Sampler model becomes more flexible, allowing for equivalent “noise” levels in the same [0, .50] range as the PT+N model. Though it would seem a more fundamental change, this same model may also be seen as a version of the PT+N theory that jettisons its constructive account of conditional probability judgment. Thus, fitting this additional “hybrid” model allows for a complete comparison of the models along both of their differing dimensions.

@@

Data from Experiment 1 are only modestly informative with respect to discriminating among the models. However, data from Experiment 2 reveal a clearer winner: a hierarchical implementation of the “hybrid” Bayesian Sampler model that removes the restriction on $\beta$, allowing it to range from [0, $\infty$] (again, lower scores for $LOO_{ic}$ are better).

>>>>

---
title             : "Comparing probabilistic accounts of probability judgments"
shorttitle        : "Comparing probability judgment accounts"

@@

abstract: |
  Bayesian theories of cognitive science hold that cognition is fundamentally probabilistic. However, one mark sometimes held against these theories is that people’s explicit probability judgments often violate the laws of probability. Two recent proposals, the "Probability Theory plus Noise" [@costello.watts2014] and "Bayesian Sampler" [@zhu.etal2020] theories of probability judgments, both seek to account for these biases while maintaining that mental credences are fundamentally probabilistic. These theories fit quite differently into the larger project of Bayesian cognitive science, but their many similarities complicate comparisons of their predictive accuracy. In particular, comparing the models demands a careful accounting of model complexity. Here, I cast these theories into a Bayesian data analysis framework that supports principled model comparison using information criteria. Compared the fits of both models on data collected by Zhu and colleagues [-@zhu.etal2020] I find the data are best explained by a version of the Bayesian Sampler model that allows for informative priors.

@@

Bayesian theories and approaches offer a unified formal framework for cognitive science [@tenenbaum.etal2011] that has had remarkable successes across domains, including in perception [e.g. @kersten.etal2004], memory [@anderson1991], language [@xu.tenenbaum2007], and reasoning [@lu.etal2012]. At the heart of the Bayesian project is the idea that cognition is fundamentally probabilistic: that people reason according to subjective degrees of belief which follow the laws of probability and, in particular, that they are revised in light of evidence according to Bayes' Rule. It is somewhat embarrassing then, that that these theories have often been accused of failing to describe human "beliefs" of the simple and everyday sort, such as beliefs like "it will rain tomorrow", "vaccines are safe," or "this politician is trustworthy".

@@

Unfortunately, people’s explicit probability judgments routinely violate the axioms of probability theory. For example, human probability judgments often exhibit the "conjunction fallacy": people will often judge the conjunction of two events (e.g. "Tom Brady likes football and miniature horses") as being more probable than one of the events in isolation (e.g. "Tom Brady likes miniature horses"), a plain and flagrant violation of probability theory [@tversky.kahneman1983]. Other demonstrations of the incoherence of probability judgments include disjunction fallacies, subadditivity or "unpacking" effects [@tversky.koehler1994], and a variety of other effects illustrating the incoherence of human probability judgments [for an accessible review, see @kahneman2013]. Altogether these findings have led many researchers to abandon the notion that degrees of belief are represented as probabilities.

Recently however, two groups of researchers have proposed theories of human probability judgments that account for biases and apparent incoherence in these judgments while maintaining that mental credences are fundamentally probabilistic [@costello.watts2014; @zhu.etal2020]. Both of these theories build on the increasingly popular notion that a variety of human reasoning tasks are accomplished by a limited process of mental "sampling" from a probabilistic mental model [see also @chater.etal2020; @dasgupta.etal2017].

@@

Costello & Watts [-@costello.watts2014; -@costello.watts2016; -@costello.watts2018] have proposed a theory of probability judgment they call the "Probability Theory plus Noise" theory (PT+N). In the PT+N model, mental "samples" are drawn from a probabilistic mental model of events and are then "read" with noise, so that some positive examples will be read as negative and some negative examples read as positive. The results are probability judgments reflecting probabilistic credences perturbed by noise. In their model, the probability a mental sample for an event $A$ is read as A is the probability that the sample truly is A, p(A), and that it is correctly read (1-d), plus the probability that the sample is not A, 1-P(A) and that it is incorrectly read (d), or:

\begin{align*}
  P(\text{read as A}) &= (1-d)P(A) + d(1-P(A)) \\
  &= (1-2d)P(A) + d
\end{align*}

Thus under the simplest form of the PT+N model, the expected value of probability judgments is:

@@

where *d* represents the probability with which samples will be misread, or the amount of noise in the judgment process. By assumption, a maximum of 50% of samples can be misread, so that d is a number in the range [0, .50]. The PT+N theory provides a unified accounts for a wide variety of biases in probability judgment that were previously attributed to different types of heuristics [@costello.watts2014; @costello.watts2016; @costello.watts2017; @costello.watts2018].

Meanwhile, Zhu, Sanborn, & Chater [-@zhu.etal2020] have proposed a Bayesian model of probability judgment they call the "Bayesian Sampler." Under this model, probability judgment is itself seen as a process of Bayesian inference. To judge the probability of an event, a limited number of samples are again drawn from a mental model of the event. Then, those "observed" samples are integrated with a prior over probabilities to produce a probability judgment. The $N$ mentally-observed samples are integrated with a Bayesian prior over probabilities, represented with a symmetric Beta distribution parameterized by $\beta$. After observing $S(A)$ successes and $N - S(A)$ failures, this posterior is distributed $Beta(\beta + S(A), \beta + N - S(A))$. Zhu and colleagues [-@zhu.etal2020] assume that people report the mean of their posterior probability estimates. Since for any Beta distribution $x \sim Beta(a,b)$, $E[x] = \frac{a}{a+b}$, the probability estimate is thus a linear function of S, N, and $\beta$.

$$\hat{P}_{BS}(A) = \frac{S(A)}{N+2\beta} + \frac{\beta}{N+2\beta}$$

The expected value of the estimate can then be written in terms of the expected number of successes, or $P(A) \cdot N$. Under the simplest version of the Bayesian Sampler model, this gives the following formula:

@@

Like the PT+N model, the Bayesian Sampler model accounts for a wide array of biases in probability judgments, including the novel biases identified by Costello and Watts [@costello.watts2014; @costello.watts2016]. In fact, important equivalencies can be drawn between the two models. Zhu and colleagues [-@zhu.etal2020] show that the $N$ and $\beta$ parameters of their model can be related to the $d$ parameter of the PT+N model via the following bridging formula:

@@

Thus, the effect of a Bayesian prior is nearly identical to the effect of noise in the PT+N model. But, rather than merely perturbing people's probability judgments, this prior can be seen as regularizing these judgments away from extreme values. Zhu and colleagues [-@zhu.etal2020] argue that such regularization can be adaptive in cases where only a small number of mental samples can be drawn. For instance, consider someone estimating the probability that they can swim across a lake, outrun an animal, or win a hand of poker: if a mental simulation of these events produces two samples indicating success, a person incorporating no other information might conclude these are all certain victories and thereby be too willing to assume risk. A regularizing prior pushes these estimates away from extremes, thereby promoting better decision-making when mental samples are sparse. However, this hedging comes at the cost of systematic incoherence and biases.

@@

By explaining the incoherence of human probability judgments using coherent mental probabilities, both models have the potential to rescue the larger project of Bayesian cognitive science as applied to everyday beliefs [@chater.etal2020]. And the bridging condition identified by Zhu and colleagues [@zhu.etal2020] demonstrates that the theories make identical predictions in many cases. However, the two models diverge substantially in their treatment of conditional probability judgments. Bayesian cognitive theories are fundamentally theories of inductive reasoning: Bayes' rule describes how existing beliefs should be updated conditional on the observation of different kinds of evidence. So, how they treat the conditioning of beliefs is at the heart of these theories.

@@

$$P_e(A|B) = \frac{(1-2d)^2P(A \land B) + d(1-2d)\big(P(A)+P(B)\big)+d^2}{(1-2d)P(B)+d}$$

@@

Zhu, Sanborn, and Chater [-@zhu.etal2020] compared their Bayesian Sampler model against Costello & Watts’ (2014; 2016; 2017; 2018) PT+N model as explanations for human probability judgments in two experiments. Unfortunately, their results were somewhat equivocal. They fit both "simple" and "complex" versions of each model and computed a Bayesian Information Criteria score (BIC) for all models. The complex versions of these models introduce additional parameters d’ and N’ that allow for different patterns of judgments for conjunctive and disjunctive judgments as compared with simple probability judgments. These additional parameters are crucial to both models’ explanations of conjunction fallacies---key findings in the probability judgment literature [@costello.watts2017; @zhu.etal2020]. Table 1 below presents the total BIC scores computed for each model as originally fit, using the authors’ original code and saved model outputs [@zhu.etal2020, Supplementary materials].

@@

In both experiments, the simple version of the Bayesian Sampler scores best, but the complex version of the PT+N model comes in second (lower BIC scores are better). It is not obvious what conclusions should be drawn from these results. Of course, the simple Bayesian Sampler model appears to win by the numbers. But conjunction fallacies are an extremely robust empirical finding [@mellers.etal2001; @sides.etal2002] and are clearly present in the data collected by Zhu and colleagues [-@zhu.etal2020]. So we might justifiably rule out the simple variants of the models on the grounds that they will fail to capture important qualitative features of the data, which would instead favor the PT+N theory. For their part, Zhu and colleagues [-@zhu.etal2020] elect to split the baby by averaging the simple and complex model scores together. This approach somewhat favors the Bayesian Sampler theory overall, though they are cautious about drawing any strong conclusions in favor of their theory.

@@

When comparing and selecting models based on their predictive accuracy, it is necessary to correct for the potential for these models to "overfit" the data. Typically this correction comes in the form of a "complexity penalty," penalizing models in proportion to their flexibility for accommodating different patterns of data [@gelman.etal2014]. Zhu and colleagues [-@zhu.etal2020] warn that BIC, which penalizes models based only on the number of parameters in the model, cannot fully account for the differences in the models’ complexity.

There are at least three challenges to accounting for model complexity in the comparison of the PT+N and Bayesian Sampler models. First, the models differ not only in the number of parameters but in the domain of those parameters. Zhu and colleagues [-@zhu.etal2020] assume that the Bayesian Sampler model’s prior distribution should reflect ignorance or a lack of information. A uniform prior, Beta(1, 1) is the most obvious choice in this case, but theoretical arguments can also be made for Beta(0.5, 0.5) (Jeffrey’s prior), Beta(0,0) (Haldene’s prior), or perhaps other  symmetric beta distributions Beta($\beta$,$\beta$) with $beta \leq 1$ [@jaynes2003]. Thus, they restrict the range of the $\beta$ parameter in their Bayesian Sampler model to fall within the interval [0,1]. Via the bridging conditions (eq X), they show that this restricts the "noise level" for the Bayesian Sampler,  represented in implied d under the PT+N model, to [0, $1/3$]. The Bayesian Sampler’s implied $d$ approaches the upper limit of $1/3$ as $N$ approaches $\infty$ and $\beta$ equals 1,  whereas the PT+N model permits noise values between [0, 0.50].

Second, it is not immediately clear how the models’ structural differences impact their flexibility. To be sure, Zhu and colleagues’ [-@zhu.etal2020] bridging condition makes it clear that the PT+N model is more flexible when it comes to predicting unconditional probabilities. But, what impact does the PT+N model’s treatment of conditional probabilities have on model complexity? Is this component of the model a sort of Ptolemaic epicycle, adding complexity to the model that should be penalized? Or, does it constitute a commitment to novel predictions that thereby constrain its flexibility? Determining model complexity a priori is not always straightforward when the models being compared differ structurally.

Finally, one potential explanation for the relative weakness of the complex model versions is Zhu and colleagues’ [-@zhu.etal2020] use of "unpooled" models, where parameters are estimated independently for each individual participant. Adding a parameter for each participant may effectively over-penalize the complex variants relative to the simple variants, especially if there is actually limited heterogeneity across participants. In contrast, a comparison of fully pooled variants of the simple and complex models (where a single population parameter is used for all participants) would require adding only one extra parameter to the penalty term. Partial pooling is a solution that balances between these extreme approaches, allowing for an accounting of heterogeneity without over-penalizing in cases where heterogeneity is low.

@@

Here, I cast both the Bayesian Sampler and PT+N models into a Bayesian data analysis framework that may permit a more decisive comparison. First, Bayesian data analysis allows issues of model complexity to be addressed through comparisons of model fit based on information criteria, such as $WAIC$ and $LOO_{ic}$. Both of these indices approximate leave-one-out cross validation, thereby offering a measure of the ability of a model to capture new unseen data. Importantly, these indices also allow for complexity penalties that directly capture differences in the flexibility of the models, rather than being estimated based solely on the number of distinct parameters [@gelman.etal2014; @vehtari.etal2017]. In addition, the Bayesian framework also supports straightforward implementation of hierarchical versions of these models with partial pooling. This allows for information about model parameters to be shared across participants, resulting in a potential reduction in model complexity and a more realistic test of the models.

@@

Zhu, Sanborn, & Chater [-@chater.etal2020] conducted two experiments to compare the PT+N and Bayesian sampler theories. These experiments asked participants to judge the probability of different events in various combinations. Following prior work by Costello and Watts [e.g. -@costello.watts2016; -@costello.watts2018], both experiments focused on the everyday events of different kinds of weather.

Experiment 1 asked about the events [icy, frosty] and [normal, typical] (e.g. "what is the probability that the weather in London is normal and not typical?").  The authors’ goal was to ask about highly correlated events, but the events used are perhaps nearly perfectly correlated. Because the terms used to describe these events are nearly synonymous, there is a concern about the interpretation of the statements evaluated in this experiment. This is especially clear, as the authors note, for disjunctive trials such as "normal or typical," where "or typical" might not be read as a disjunction but rather an elaborative clause.  In light of these concerns, I excluded the disjunctive trials from Experiment 1 from my analyses.

Experiment 2 focused on more moderately correlated events, [cold, rainy] and [windy, cloudy], that do not admit these misinterpretations. In addition, a third experimental condition asking about [warm, snowy] was also included in the experiment, but was dropped from the analyses reported in the paper. Exploring the raw responses from this condition reveals a substantial fraction of "zero" and "one" responses for certain trials. This may reflect a different response process than was intended. For instance, some participants may have engaged in deductive reasoning to judge that it is not possible for the weather to at once be warm and snowy, and therefore responded with zero—failing to properly consider that it is possible (at least logically) for it to be warm and snowy at different times within the same day. Given these potentially aberrant responses, I followed Zhu and colleagues [-@zhu.etal2020] in ignoring data from this condition.

@@

The PT+N model defines probability judgments as:

@@

In contrast, the Bayesian Sampler model defines probability judgments as:

@@

Fixing d and d’ or N and N’ equal yields the "simple" variant of each of the models, which treat conjunctive and disjunctive probability judgments identically to simple probability judgments.

Notice that for each model the probability judgments depend on underlying subjective probabilities, derived from a mental sampling process. These subjective probabilities are unobserved, and must be estimated as a latent variable. Here, they are represented with a four-dimensional dirichlet distribution for each subject, representing the probability of the elementary events ($A \land B$, $\neg A \land B$, $A \land \neg B$, $\neg A \land \neg B$).

Zhu, Sanborn & Chater (2020) implement completely unpooled models with separate d, d’, N, N’, and $\beta$ parameters for each participant. Figure 1 displays the translation of the PT+N model into the Bayesian framework, along with a plate diagram representing the dependencies among parameters.

\begin{figure}
  \begin{minipage}{.5\textwidth}
    \flushright
    \input{ptn_platediagram}
  \end{minipage}%
  \begin{minipage}{.5\textwidth}
    \flushleft
    \small
    \begin{align*}
      y_{ijk} &\sim beta\big(\mu_{ijk} \kappa, (1-\mu_{ijk}) \kappa \big) \\
      \mu_{ijk} &= f_{PT+N}(\overrightarrow{p_{jk}}, d_j, d_j)  \\
      d_j &= \frac{expit(\delta_j)}{2} \\
      d_j’ &= \frac{expit\big(\delta_j + \exp(\Delta\delta_j)\big)}{2} \\
      \overrightarrow{p_{jk}} &\sim Dirichlet(\overrightarrow{1}) \\
      \delta_j &\sim Normal(0,1) \\
      \Delta\delta_j &\sim Normal(0,1) \\
      \kappa &\sim HalfCauchy(20)
    \end{align*}
  \end{minipage}
  \caption{Complex unpooled PT+N model diagram and formula specifications. Circular nodes are parameters, shaded nodes are observations, and squared nodes are deterministic functions of parameters. Plates signify values defined for $i$ trials, $j$ participants, and $k$ conditions.}
\end{figure}

@@

Recall that Zhu and colleagues (2020) identified a bridging condition relating b and N in the Bayesian Sampler model to the *d* parameter of the PT+N model. To support direct comparisons of the models, I parameterize the Bayesian Sampler model according to the implied d and d', rather than directly according to its $\beta$, N, and N' parameters. I constrain *d* to $[0, 1/3]$ for the Bayesian Sampler model to reflect the assumption that $\beta \in [0, 1]$. This allows the same priors to be used for the corresponding Bayesian Sampler and PT+N models, simplifying their comparison.

The Bayesian Sampler model is therefore identical to the PT+N model save for thechanges to $\mu_{ijk}$, $d$, and $d'$ shown below:

\begin{align*}
  \mu_{ijk} &= f_{BS}(\overrightarrow{p_{jk}}, d_j, d_j)  \\
  d_j &= \frac{expit(\delta_j)}{3} \\
  d_j’ &= \frac{expit\big(\delta_j + \exp(\Delta\delta_j)\big)}{3} \\
\end{align*}

@@

Both of these models can also be implemented as hierarchical models with partial pooling for the d and d’ parameters (implicitly, for N and N’ in the case of the Bayesian Sampler). This should prevent over-penalizing complex versions of the models and potentially improve their overall predictive fit. The hierarchical implementation adds parameters for the population-level d and d’ as well as a parameter controlling the standard deviation of the distribution for the subject-level effects. For ease of interpretation, the centered parameterization is shown below, although the actual models used a non-centered parameterization to improve sampling efficiency [@papaspiliopoulos.etal2007]. Figure 2 displays the translation of a hierarchical implementation of the Bayesian Sampler model into the Bayesian framework, along with a plate diagram representing the dependencies among parameters.

\begin{figure}
  \begin{minipage}{.5\textwidth}
    \flushright
    \input{bs_platediagram_mlm}
  \end{minipage}%
  \begin{minipage}{.5\textwidth}
    \flushleft
    \small
    \input{bs_equation_mlm}
  \end{minipage}
  \caption{Hierarchical complex Bayesian Sampler model diagram and formula specifications. Circular nodes are parameters, shaded nodes are observations, and squared nodes are deterministic functions of parameters. Plates signify values defined for $i$ trials, $j$ participants, and $k$ conditions.}
\end{figure}

Finally, I also explored fitting versions of the Bayesian Sampler model that allowed values of $\beta > 1$. Restricting $\beta$ to [0,1] restricts the prior distribution of the Bayesian sampler to the class of "ignorance priors" [@zhu.etal2020]. However, it is also possible that people bring informative priors to the probability judgment task. Indeed, Zhu and colleagues (2020) acknowledge there are situations where an informative prior may be warranted [see e.g., @fennell.baddeley2012]. If $\beta$ is unrestricted, allowed to fall in the domain [0, $\infty$] then the Bayesian Sampler model becomes more flexible, allowing for equivalent "noise" levels in the same [0, .50] range as the PT+N model. That is, through the bridging condition, the implied $d$ approaches .50 in the limit as $N$ and $\beta$ both approach $\infty$. Though it would seem a more fundamental change, this same model may also be seen as a version of the PT+N theory that jettisons its constructive account of conditional probability judgment. Thus, fitting this additional unrestricted model allows for a complete comparison of the models along both of their differing dimensions.

@@

Data from Experiment 1 are only modestly informative with respect to discriminating among the models. However, data from Experiment 2 reveal a clearer winner: a hierarchical implementation of the Bayesian Sampler model that removes the restriction on $\beta$, allowing it to range from [0, $\infty$] (again, lower scores for $LOO_{ic}$ are better).

Recall, this model may also be seen as a version of the PT+N theory that removes the original theory’s constructive account of conditional probability judgments. Comparing predictions for conditional probability judgments from the unrestricted Bayesian Sampler model and the best-fitting PT+N model, we see that the Bayesian Sampler model better captures these judgments from Experiment 2 (trial-average-level $R^2$ = .85 vs .73; participant-level $R^2$ = .51 vs .45). These findings suggest that conditioning is better seen as part of the mental model than the probability judgment process.
